---
title: "Pacific High Seas Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pacific High Seas Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This example walks though how to do a spatial prioritization using `oceandatr` to obtain suitable data, and also using the package [`patchwise`](https://github.com/emlab-ucsb/patchwise) to ensure that entire seamounts are included in the prioritization solutions.

We use a High Seas area of the Pacific as the planning area for this example since it is outside any states' jurisdiction.

```{r, include = FALSE}
#for cropping whitespace from around figures, use crop = TRUE
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

knitr::opts_chunk$set(
  echo = TRUE,
	message = FALSE,
	warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5, 
  crop = TRUE
)
```

```{r eval=FALSE}
#load oceandatr package
library(oceandatr)
```

```{r include=FALSE}
devtools::load_all()
```

Along with `oceandatr` we will need to load `prioritizr` for the spatial prioritization and we will use the open source solver `lsymphony` to solve the prioritization problem, so this needs to be installed and loaded via the Bioconductor website 

```{r setup}
library(gfwr)
library(prioritizr)
#remotes::install_bioc("lpsymphony")
library(lpsymphony)
#remotes::install_github("emlab-ucsb/patchwise")
library(patchwise)
```

# High Seas area of the Pacific Ocean

First we retrieve geospatial data for the North Pacific Ocean using `get_boundary()`, then we will crop for the area we are interested in, the planning region, highlighted in red on the map.

```{r high_seas_area, fig.cap="Map of the High Seas"}
high_seas <- get_boundary(type = "high_seas")

pacific_hs_area_extent <- sf::st_bbox(c(xmin = 135, xmax = 155, ymin = 0, ymax = 6), crs = 4326)

plot(sf::st_geometry(high_seas), main = "High Seas", col = "royalblue3", axes = TRUE, las = 1)
plot(pacific_hs_area_extent %>% sf::st_as_sfc() %>% sf::st_cast(to = "LINESTRING"), col = "red", lwd=2, add = TRUE)
```

We are going to use only the highlighted Pacific area which borders Indonesia, Papua New Guinea, Palau and the Federated States of Micronesia. We can get the EEZs of these states using `oceandatr`'s `get_area` function.

```{r get_surrounding_eezs}

country_names <- c("Indonesia", "Papua New Guinea", "Palau", "Micronesia")

eezs <- lapply(country_names, FUN = function(x) get_boundary(name = x) %>% dplyr::select(territory1) %>% dplyr::rename(name = territory1)) %>% 
  do.call(rbind, .) %>% 
  sf::st_cast(to = "MULTIPOLYGON")
```

We only want the high seas portion of the area outlined above. We can create a polygon for this planning region by removing the bordering states EEZs.

```{r pacific_high_seas}
sf::sf_use_s2(FALSE) #turn off S2 to avoid errors

pacific_hs <- high_seas %>%
  sf::st_crop(pacific_hs_area_extent) %>% 
  sf::st_sf() %>% 
  dplyr::mutate(name = "Planning region") %>% 
  dplyr::select(name)

sf::sf_use_s2(TRUE)
```

To make a nice context map, we can download bathymetry data using `oceandatr`'s `get_bathymetry()` function, using the bounding box of the EEZs as the input extent (grid). To get country boundaries to add to the map, we use the `get_boundary()` functions from `oceandatr`, setting `name = NULL` so that all countries are downloaded. We can then plot everything using the `tmap` package, which is great for making nice maps.

```{r pacific_high_sea_maps, fig.cap="Map of Pacific High Seas planning region in regional context"}

#retrieve bathymetry for the planning regions and surrounding area bounded by the EEZs
pacific_bathy <- get_bathymetry(spatial_grid = sf::st_as_sfc(sf::st_bbox(eezs), crs = 4326) %>% sf::st_as_sf(), raw = TRUE, classify_bathymetry = FALSE, resolution = 1) %>% 
  terra::classify(matrix(c(0, Inf, NA), ncol = 3))

#retrieve country boundaries using oceandatr
world <- get_boundary(name = NULL, type = "country")

tmap::tm_shape(pacific_bathy * -1) + #multiply bathymetry by -1 to make values positive for nicer legend
  tmap::tm_raster(
    col.scale = tmap::tm_scale_continuous(values = "brewer.yl_gn_bu"),
    col.legend = tmap::tm_legend(title = "Depth (m)", position = tmap::tm_pos_out("right"))
  ) +
  tmap::tm_shape(eezs %>% sf::st_cast(to = "MULTILINESTRING")) +
  tmap::tm_lines(
    col = "name",
    lwd = 1.5,
    col.scale = tmap::tm_scale_categorical(values = "brewer.dark2"),
    col.legend = tmap::tm_legend(title = "Select EEZs")
  ) +
  tmap::tm_shape(pacific_hs %>% sf::st_cast(to = "MULTILINESTRING")) +
  tmap::tm_lines(
    col = "name",
    lwd = 2,
    col.scale = tmap::tm_scale_categorical(values = "red"),
    col.legend = tmap::tm_legend(title = "")
  ) +
  tmap::tm_shape(world) +
  tmap::tm_borders() +
  tmap::tm_graticules(lwd = 0.5) +
  tmap::tm_layout(inner.margins = 0)

tmap::tmap_save(test, "../../../Downloads/case_study.png")
```


```{r include=FALSE, eval=FALSE}
#simple plot of planning region and surrounding EEZs
plot(rbind(sf::st_cast(pacific_hs, to = "MULTIPOLYGON"), eezs), axes = TRUE, main = NULL, key.pos = 4)
```


Now we select a suitable projection for the area and a suitable resolution for the planning grid used for gridding the data. We can use [projection wizard](https://projectionwizard.org) to find an equal-area projection, entering the same extent coordinates we used to crop the high seas area (xmin = 135, xmax = 155, ymin = 0, ymax = 6). 

We will use 10km square planning units so that there the data processing and prioritization run reasonably fast (smaller planning units will require more time/ computer memory get data for)

```{r planning_grid, fig.cap="Pacific High Seas area planning grid raster"}

pacific_hs_projection <- "+proj=cea +lon_0=145 +lat_ts=3 +datum=WGS84 +units=m +no_defs"

pacific_hs_planning_grid <- get_grid(boundary = pacific_hs,
                                     crs = pacific_hs_projection,
                                     resolution = 10000) 

#get_grid returns a raster by default, so we can plot it using the terra package. We convert it to polygons so we can see each grid square.
terra::plot(terra::as.polygons(pacific_hs_planning_grid, aggregate = FALSE))
```

Now we have a planning grid, we can get data on conservation features (e.g. habitats) using `oceandatr` to use in a spatial prioritization with a single command `get_features()`. We have to set the seamount buffer, which is the area around the seamount that is included as part of the seamount, and we use 30km based since biodiversity is known to be higher within this distance of seamount peaks (see `?get_seamounts_buffered` for more info).

```{r features, fig.cap="Conservation features for the Pacific High Seas planning region"}
#set seed for reproducibility in the get_enviro_zones() sampling to find optimal cluster number
set.seed(500)

feature_set <- get_features(spatial_grid = pacific_hs_planning_grid, seamount_buffer = 30000) %>% 
  oceandatr:::remove_empty_layers() %>%  #use this to remove raster layers that are empty
  terra::subst(NA, 1) %>%
  terra::mask(pacific_hs)

terra::plot(feature_set, 
            col = c("grey60", "royalblue"), 
            maxnl = terra::nlyr(feature_set),
            axes = FALSE,
            fun = function(x)terra::lines(pacific_hs %>% sf::st_transform(pacific_hs_projection))) #set maximum number of layers to plot to the same as the number of layers in the feature set
```

```{r }
#eval=FALSE, include=FALSE
#nicer maps for paper
.simpleCap <- function(x) {
    s <- strsplit(x, " ")[[1]]
    paste(toupper(substring(s, 1, 1)), substring(s, 2),
          sep = "", collapse = " ")
}

tmap::tm_shape(setNames(feature_set2, gsub("_", " ", names(feature_set)) %>% stringr::str_to_sentence())) +
  tmap::tm_raster(col.scale = tmap::tm_scale_categorical(values = c("grey70", "royalblue"),
                                                      labels = c("Absent", "Present")),
                  col.legend = tmap::tm_legend(show = FALSE)) +
  tmap::tm_shape(pacific_hs %>% sf::st_transform(pacific_hs_projection)) +
  tmap::tm_borders() +
  tmap::tm_layout(panel.label.size = 1.5)


```


# Cost data: Global Fishing Watch data

The other piece of data needed for a spatial prioritization is cost. In terrestrial spatial planning, this can be the actually monetary value of buying the land for conservation. In marine spatial planning, measures of fishing, such as catch and fishing effort, are often used as the opportunity cost for each planning unit.

[Global Fishing Watch](https://globalfishingwatch.org/) has global fishing effort data, and this can be accessed easily using `get_gfw()` function in `oceandatr` (which is a wrapper for the `get_raster()` function from the [`gfwr`](https://github.com/GlobalFishingWatch/gfwr) package). An API key is required, but can be easily generated at no cost; see the `gfwr` website for more details.

```{r fishing-effort, fig.cap="Map of total apparent fishing effort in 2022 for the Pacific High Seas area. Data from Global Fishing Watch", message=FALSE}

fishing_effort <- get_gfw(spatial_grid = pacific_hs_planning_grid, start_year = 2022, end_year = 2022, summarise = "total_annual_effort") %>% 
  terra::subst(NA, 0.01) %>% #set NA values to zero otherwise they will be left out of the prioritization
  terra::mask(pacific_hs_planning_grid) %>% 
  setNames("fishing_effort")

terra::plot(fishing_effort, fun = terra::lines(pacific_hs %>% sf::st_transform(pacific_hs_projection)), axes = FALSE)
```

```{r eval=FALSE, include=FALSE}
#pretty map of GFW data for paper
(fishing_effort_map <- tmap::tm_shape(fishing_effort) +
  tmap::tm_raster(palette = "viridis", style = "cont", title = "Total fishing\neffort 2022", legend.reverse = TRUE) +
  tmap::tm_shape(pacific_hs %>% sf::st_transform(pacific_hs_projection)) +
  tmap::tm_borders() +
  tmap::tm_layout(legend.outside = TRUE))

tmap_save(fishing_effort_map, "fishing_effort.png")
```


# Run a simple spatial prioritization

We now have all the data we need to create a conservation problem and solve it to get a map of priority areas for conservation for our Pacific High Seas area. For the prioritization, we need to set targets for how much of each conservation feature must be included in the prioritized areas. We will set this at 20%.

```{r prioritization, fig.cap="Prioritization solution for the Pacific High Seas area. Black outlines are seamounts."}
prob <- prioritizr::problem(x = fishing_effort, features = feature_set) %>% 
  add_min_set_objective() %>% 
  add_relative_targets(0.2) %>% 
  #add_boundary_penalties(penalty = 0.00001) %>% 
  add_binary_decisions() %>% 
  add_lpsymphony_solver(verbose = FALSE)

sol <- solve(prob)

terra::plot(sol, main = "Solution",  
            col = c("grey70", "green4"),
            type = "classes",
            levels = c("Not selected", "Selected"),
            fun = terra::lines(pacific_hs %>% sf::st_transform(pacific_hs_projection)),
            axes = FALSE, 
            mar = c(3,3,2,6)+0.1)
terra::plot(terra::as.polygons(feature_set[["seamounts"]]), add=TRUE)
```

```{r eval=FALSE, include=FALSE}
#pretty map of prioritization solution for paper

(prioritization_map <- tmap::tm_shape(sol) +
  tmap::tm_raster(palette = c("grey70", "green4"),style = "cat", labels =  c("Not selected", "Selected"), title = "") +
  tmap::tm_shape(pacific_hs %>% sf::st_transform(pacific_hs_projection)) +
  tmap::tm_borders() +
  tmap::tm_layout(legend.outside = TRUE))

tmap_save(prioritization_map, "prioritization.png")

tmap::tm_shape(sol) +
  tmap::tm_raster(palette = c("grey70", "green4"),style = "cat", legend.show = FALSE) +
  tmap::tm_shape(pacific_hs %>% sf::st_transform(pacific_hs_projection)) +
  tmap::tm_borders() +
  tmap::tm_layout(frame = FALSE)
```


# Prioritization with patches

To ensure that whole seamount areas area included in the solution, we need to use the `patchwise` package to do some pre-processing of the data we use. The prioritization result is similar to that above, but whole seamount patches equal to at least 20% of the total seamount area are included.

```{r prioritization_patches, fig.cap="Prioritization solution for the Pacific High Seas area, with whole seamounts included. Black outlines are seamounts."}
# Separate seamount data - we want to protect entire patches
seamounts_rast <- feature_set[["seamounts"]]
features_rast <- feature_set[[names(feature_set)[names(feature_set) != "seamounts"]]]

# Create seamount patches - seamount areas that touch are considered the same patch
patches_rast <- patchwise::create_patches(seamounts_rast %>% terra::subst(0, NA)) #patchwise currently expects all non-patches to be NA, some subsituting non-seamounts, which are currently zeroes, with NAs

# Create patches dataframe - this creates several constraints so that entire seamount units are protected together
patches_df_rast <- patchwise::create_patch_df(spatial_grid = pacific_hs_planning_grid, features = features_rast, patches = patches_rast, costs = fishing_effort)

# Create boundary matrix for prioritizr
boundary_matrix_rast <- patchwise::create_boundary_matrix(spatial_grid = pacific_hs_planning_grid, patches = patches_rast, patch_df = patches_df_rast)

# Create targets for protection - let's just do 20% for each feature (including 20% of whole seamounts)
targets_rast <- patchwise::features_targets(targets = rep(0.2, (terra::nlyr(features_rast) + 1)), features = features_rast, pre_patches = seamounts_rast)

# Add these targets to targets for protection for the "constraints" we introduced to protect entire seamount units
constraints_rast <- patchwise::constraints_targets(feature_targets = targets_rast, patch_df = patches_df_rast)

# Run the prioritization
problem_rast <- prioritizr::problem(x = patches_df_rast, features = constraints_rast$feature, cost_column = "fishing_effort") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_manual_targets(constraints_rast) %>%
  prioritizr::add_binary_decisions() %>%
  #prioritizr::add_boundary_penalties(penalty = 0.0001, data = boundary_matrix_rast) %>%
  prioritizr::add_lpsymphony_solver()

# Solve the prioritization
solution_rast <- solve(problem_rast)

# Convert the prioritization into a more digestible format
result_rast <- patchwise::convert_solution(solution = solution_rast, patch_df = patches_df_rast, spatial_grid = pacific_hs_planning_grid)

# Show the results
terra::plot(result_rast, main = "Solution", 
            col = c("grey70", "green4"),
            type = "classes",
            levels = c("Not selected", "Selected"),
            fun = terra::lines(pacific_hs %>% sf::st_transform(pacific_hs_projection)), 
            axes = FALSE,
            mar = c(3,3,2,6)+0.1)
terra::plot(terra::as.polygons(seamounts_rast), add=TRUE)
```

```{r eval = FALSE, include=FALSE}
#nice maps for publication

# tmap::tm_shape(total_annual_effort) +
#   tmap::tm_raster(palette = "viridis", title = "Total apparent fishing\neffort (hours)", style = "quantile", n = 8) +
#   tmap::tm_layout(legend.outside = TRUE)

ggplot_gfw <- ggplot2::ggplot() +
  ggplot2::geom_raster(data = terra::as.data.frame(total_annual_effort %>% terra::project("epsg:4326"), xy = TRUE), ggplot2::aes(x=x, y=y, fill = fishing_effort)) +
  ggplot2::coord_sf() +
  ggplot2::scale_fill_viridis_c(name = "Apparent fishing\neffort (hours)") +
  ggplot2::scale_y_continuous(labels = ~ paste0(.x, "°N")) +
  ggplot2::scale_x_continuous(labels = ~ paste0(.x, "°E")) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.title = ggplot2::element_blank())

ggplot_sol_no_patches <- ggplot2::ggplot() +
  ggplot2::geom_raster(data = terra::as.data.frame(sol %>% terra::project("epsg:4326", method = "near"), xy = TRUE), ggplot2::aes(x=x, y=y, fill = factor(fishing_effort))) +
  ggplot2::geom_sf(data = seamounts_rast %>% terra::as.polygons() %>% terra::project("epsg:4326") %>% sf::st_as_sf() %>% dplyr::mutate(type = "Seamounts"), ggplot2::aes(colour = type), fill = NA, lwd = 0.2) +
  ggplot2::coord_sf() +
  ggplot2::scale_fill_manual(values = c("grey", "forestgreen"),
                             labels = c("Not selected", "Selected"),
                             name = NULL) +
  ggplot2::scale_color_manual(values = c("Seamounts" = "black"),
                              labels = "Seamounts",
                              name = NULL) +
  ggplot2::scale_y_continuous(labels = ~ paste0(.x, "°N")) +
  ggplot2::scale_x_continuous(labels = ~ paste0(.x, "°E")) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.title = ggplot2::element_blank())

ggplot_sol_patches <- ggplot2::ggplot() +
  ggplot2::geom_raster(data = terra::as.data.frame(result_rast %>% terra::project("epsg:4326", method = "near"), xy = TRUE), ggplot2::aes(x=x, y=y, fill = factor(layer))) +
    ggplot2::geom_sf(data = seamounts_rast %>% terra::as.polygons() %>% terra::project("epsg:4326") %>% sf::st_as_sf() %>% dplyr::mutate(type = "Seamounts"), ggplot2::aes(colour = type), fill = NA, lwd = 0.2) +
  ggplot2::coord_sf() +
  ggplot2::scale_fill_manual(values = c("grey", "forestgreen"),
                             labels = c("Not selected", "Selected"),
                             name = NULL) +
    ggplot2::scale_color_manual(values = c("Seamounts" = "black"),
                              labels = "Seamounts",
                              name = NULL) +
  ggplot2::scale_y_continuous(labels = ~ paste0(.x, "°N")) +
  ggplot2::scale_x_continuous(labels = ~ paste0(.x, "°E")) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.title = ggplot2::element_blank())


legend_gfw <- cowplot::get_legend(
  # create some space to the left of the legend
  ggplot_gfw + ggplot2::theme(legend.box.margin = ggplot2::margin(0, 0, 0, 0))
)

legend_sols <- cowplot::get_legend(
  # create some space to the left of the legend
  ggplot_sol_no_patches + ggplot2::theme(legend.box.margin = ggplot2::margin(0, 0, 0, 0))
)

plots <- cowplot::plot_grid(ggplot_gfw+ ggplot2::theme(legend.position = "none", plot.margin = ggplot2::margin(0,0,6,0)), ggplot_sol_no_patches + ggplot2::theme(legend.position = "none", plot.margin = ggplot2::margin(0,0,6,0)), ggplot_sol_patches + ggplot2::theme(legend.position = "none", plot.margin = ggplot2::margin(0,0,6,0)), ncol = 1, labels = "auto", label_size = 12)

legends <- cowplot::plot_grid(legend_gfw, legend_sols, ncol = 1, rel_heights = c(0.4, 1))

cowplot::plot_grid(plots, legends, ncol = 2, rel_widths = c(0.8, 0.2))
```

